\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-11-08 metatheory-condensed.tex
executive summary of metatheory

We contribute \TT, a core type theory that incorporates the core elements
of \underline{f}a\underline{m}i\underline{l}y polymorphism into
\underline{M}artin-\underline{L}öf dependent \underline{t}ype \underline{t}heory~\cite{martin1982constructive},
while maintaining consistency and canonicity.\NA{maybe you can give an overview of the salient points of the design and what they're trying to achieve? What is the core aspect modeled by the theory?}

Most notably, \TT extends MLTT with what we call \emph{linkages}
(a namesake of the theoretical device through which \citet{zm2017} model
family polymorphism in an OO setting).
Linkages model families, so they are like tuples but with a twist: they
support late binding.

\TT is intended as a foundational model. So unlike our programmer-facing plugin,
typing is structural rather than nominal, and
it supports family polymorphism by providing MLTT-style typing rules that
can be used to \emph{encode} core aspects of the language supported by our
plugin, rather than building them in directly.
%
For example, \TT supports extending inductive types through
formation rules that enable the encoding of the further binding of W-types~\cite{martin1984intuitionistic} in linkages.
Also unlike our plugin, \TT does not feature automations such as
the generation of propositional equalities and partial recursors,
but it is expressive enough to encode their definitions.

@@

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT, and
\cref{fig:fmltt-selected} \TT.
%The new forms (relative to MLTT) are highlighted.
We briefly review the base MLTT fragment first.

As is standard in MLTT presentations, we use explicit substitutions~\cite{substcalculus,abadi1989subst}:
substitutions~$\gm$ and their applications (e.g., \sub{T}{\gm}) are part
of the syntax rather than meta-operations.
%
Variables are represented by de Bruijn indices:
\var{n} is the variable bound by the $n$-th closest enclosing binder.
For example, $\lambda x.\,\lambda y.\,x$ is $\lam{\lam{\var{1}}}$.
%
Typing judgments for substitutions have the form 
%\EDJ{It seems to me G |- g : D is a judgement not a rule. Rules usually have two parts (premises + conclusion). Judgment is a standard terminology invented by Martin Lof (https://ncatlab.org/nlab/show/judgment see reference here). I will suggest saying 'Formation rules for substitution lead to a judgement G |- g : D' sth like this. In pen-and-paper, the formation rule (corresponding to QIIT well-typed-ness of judgement) is called presuppose (can be found in Modal MLTT paper) }
$\goodSub{\Gm}{\gm}{\Dl}$.
The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
yields terms valid in~$\Gm$ (e.g., rule \ruleref{tm/sub}).
%
The two main forms of substitutions are weakening (\ruleref{sub/wk}) and extension (\ruleref{sub/ext}):
\sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
\sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
then applies~$\gm$.
For example, rule \ruleref{tm/snd} states that if $t$ is a dependent pair
that has type $\TySigma{A}{B}$, then $\snd{t}$ has type
$\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$.

>>>>

We contribute \TT, a core type theory that extends
%\underline{f}a\underline{m}i\underline{l}y polymorphism
%\underline{M}artin-\underline{L}öf dependent \underline{t}ype \underline{t}heory (MLTT)~\cite{martin1982constructive}
Martin-Löf dependent type theory (MLTT)~\cite{martin1982constructive}
with facilities to express family polymorphism
while maintaining consistency and canonicity.
%\NA{maybe you can give an overview of the salient points of the design and what they're trying to achieve? What is the core aspect modeled by the theory?}
%
We acknowledge that the presentation herein is dense for an audience
without prior exposure to MLTT, so we summarize the salient points
first before delving into the technical details.

\noindentparagraph{Summary.}

\TT is intended as a foundational model. So unlike our programmer-facing plugin,
typing is structural rather than nominal, and
partial recursors and propositional equalities are not automated but have to
be explicitly defined.
\TT provides MLTT-style constructs that can be used to \emph{encode}
family polymorphism rather than building it in directly.
%
Most notably, \TT extends MLTT with what we call \emph{linkages}.\footnote{
Linkages are a namesake of the theoretical device through which prior work~\cite{zm2017} models
family polymorphism in an OO setting, but the technical details differ significantly.}
Linkages model families, so they are like tuples but with a twist: they
support late binding.
Unlike dependent tuples where a later component is \emph{existentially}
quantified over the earlier ones, a linkage component is \emph{universally}
quantified over---and thus polymorphic to---components preceding it.
%
\TT also features \emph{linkage transformers} that can be used to inductively
construct linkages from other linkages, by inheriting or overriding components of a linkage,
or by extending linkages with new components.
%
Inductive types are modeled as W-types~\cite{martin1984intuitionistic} and their
extension as overriding.
%
Consistency and canonicity are proved by giving semantic models to the syntax.

%\TT supports extending inductive types through
%formation rules that enable the encoding of the further binding of W-types~\cite{martin1984intuitionistic} in linkages.

%Also unlike our plugin, \TT does not feature automations such as
%the generation of propositional equalities and partial recursors,
%but it is expressive enough to encode their definitions.

@@

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT, and
\cref{fig:fmltt-selected} \TT.
%The new forms (relative to MLTT) are highlighted.
We briefly review the base MLTT fragment first.
%
As is standard in MLTT presentations, we use explicit substitutions~\cite{substcalculus,abadi1989subst}:
substitutions~$\gm$ and their applications (e.g., \sub{T}{\gm}) are part
of the syntax rather than meta-operations.
%
Variables are represented by de Bruijn indices:
\var{n} is the variable bound by the $n$-th closest enclosing binder.
For example, $\lambda x.\,\lambda y.\,x$ is $\lam{\lam{\var{1}}}$.
%
Typing judgments for substitutions have the form 
%\EDJ{It seems to me G |- g : D is a judgement not a rule. Rules usually have two parts (premises + conclusion). Judgment is a standard terminology invented by Martin Lof (https://ncatlab.org/nlab/show/judgment see reference here). I will suggest saying 'Formation rules for substitution lead to a judgement G |- g : D' sth like this. In pen-and-paper, the formation rule (corresponding to QIIT well-typed-ness of judgement) is called presuppose (can be found in Modal MLTT paper) }
$\goodSub{\Gm}{\gm}{\Dl}$.
The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
yields terms valid in~$\Gm$ (e.g., rule \ruleref{tm/sub}).
%
The two main forms of substitutions are weakening (\ruleref{sub/wk}) and extension (\ruleref{sub/ext}):
\sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
\sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
then applies~$\gm$.
For example, rule \ruleref{tm/snd} states that if $t$ is a dependent pair
that has type $\TySigma{A}{B}$, then $\snd{t}$ has type
$\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$.
